\pkg{partykit}: A Modular Toolkit for Recursive Partytioning in \proglang{R}

### 2014-03-25 pkg/partykit/inst/mloss/mloss2.Rnw
touch-ups in sections 1-3
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@553 255f4249-2263-4417-b1ea-8498779e27a7

%% software markup
\let\proglang=\textsf
\let\pkg=\textit
\let\code=\texttt
\newcommand{\squote}[1]{`{#1}'}
\newcommand{\dquote}[1]{``{#1}''}
\newcommand{\fct}[1]{\texttt{#1()}}
\newcommand{\class}[1]{\squote{\texttt{#1}}}

@@

\begin{abstract}
  The \pkg{partykit} package provides a flexible toolkit with infrastructure for
  learning, representing, summarizing, and visualizing a wide range of tree-structured
  regression and classification models. The functionality encompasses: (a)~Basic
  infrastructure for \emph{representing} trees (inferred by any algorithm) so that
  unified \code{print}/\code{plot}/\code{predict} methods are available.
  (b)~Dedicated methods for trees with \emph{constant fits} in the leaves (or terminal nodes)
  along with suitable coercion functions to create such tree models (e.g., by
  \pkg{rpart}, \pkg{RWeka}, PMML). (c)~A reimplementation of \emph{conditional inference
  trees} (\code{ctree}, originally provided in the \pkg{party} package). (d)~An extended
  reimplementation of \emph{model-based recursive partitioning} (\code{mob}, also originally
  in \pkg{party}) along with dedicated methods for trees with parametric models in the
  leaves. Here, a brief overview of the package and its design is given while more detailed
  discussions of items~(a)--(d) are available in vignettes accompanying the package.
\end{abstract}

@@

In the more than fifty years since \cite{Morgan+Sonquist:1963}
published their seminal paper on ``automatic interaction detection'',
a wide range of methods has been suggested that is usually
termed ``recursive partitioning'' or ``decision trees'' or
``tree(-structured) models'' etc. Particularly influential were the
algorithms CART \citep[classification and regression trees,][]{Breiman+Friedman+Olshen:1984},
C4.5 \citep{Quinlan:1993}, QUEST/GUIDE \citep{Loh+Shih:1997,Loh:2002},
and CTree \citep{Hothorn+Hornik+Zeileis:2006} among many others
\citep[see][for a recent overview]{Loh:2014}. Reflecting the heterogeneity
of conceptual algorithms, a wide range of computational implementations
in various software systems emerged: Typically the original authors of an
algorithms also provide accompanying software but many software systems,
e.g., including \pkg{Weka} \citep{Witten+Frank:2005} or \proglang{R} \citep{R},
also provide collections of various types of trees.
Within \proglang{R} the list of prominent packages includes
\pkg{rpart} \citep[implementing the CART algorithm]{rpart},
\pkg{RWeka} \citep[containing interfaces to J4.8, M5', LMT from \pkg{Weka}]{RWeka}, and
\pkg{party} \citep[implementing CTree and MOB]{party} among
many others. See the CRAN task view ``Machine Learning''
\citep{ctv} for an overview.

@@

The \pkg{partykit} tries to address the latter point and improve the
computational situation by providing a common unified infrastructure
for recursive partytioning in the \proglang{R} system for statistical
computing. In particular, \pkg{partykit}
provides tools for representing fitted trees along with printing,
plotting, and computing predictions. The design principles are:
\begin{itemize}
  \item One `agnostic' base class (\class{party}) which can encompass an extremely wide
        range of different types of trees.
  \item Subclasses for important types of trees, e.g., trees with constant
        fits (\class{constparty}) or with parametric models (\class{modelparty})
	in each terminal node (or leaf).
  \item Nodes are recursive objects, i.e., a node can contain child nodes.
  \item Keep (learning) data out of the recursive node and split structure.
  \item Basic printing, plotting, and predicting for raw node structure.
  \item Customization via suitable panel or panel-generating functions.
  \item Coercion from existing objects (\code{rpart}, \code{J48}, etc.) to the new class.
  \item Usage of simple/fast \proglang{S}3 classes and methods.
\end{itemize}
In addition to all of this generic infrastructure, two specific tree
algorithms are implemented in \pkg{partykit} as well: \fct{ctree}
for conditional inference trees \citep{Hothorn+Hornik+Zeileis:2006}
and \fct{mob} for model-based recursive partitioning \citep{Zeileis+Hothorn+Hornik:2008}.

@@

The \pkg{partykit} package is an add-on package for the \proglang{R} system
for statistical computing. It is available from the Comprehensive \proglang{R}
Archive Network (CRAN) at \url{http://CRAN.R-project.org/package=partykit} and
can be easily installed in the usual ways from within \proglang{R} using
\fct{install.packages}. It depends on \proglang{R} (at least 2.15.0) as well
as the base packages \pkg{graphics}, \pkg{grid}, \pkg{stats}, and the recommended
\pkg{survival} package. Furthermore, various suggested packages are needed
for certain special functionalities in the package. To install all of these
required and suggested packages in one go, the command
\code{install.packages("partykit", dependencies = TRUE)} can be used.

In addition to the stable release version of the package on CRAN, the
current development release can be installed from \proglang{R}-Forge
\citep{Theussl+Zeileis:2009}. In addition to source and binary packages
of \pkg{partykit} the entire version history is available through
\proglang{R}-Forge's \pkg{Subversion} source code management system.

Along with the package extensive documentation and examples are shipped.
The manual pages provide basic technical information on all functions while
much more detailed descriptions along with worked examples are provided in
the four package vignettes. First, the vignette \code{"partykit"}
introduces the basic \class{party} class
and associated infrastructure while three further vignettes discuss the tools
built on top of it: \code{"constparty"} covers the eponymous class (as well as
the simplified \class{simpleparty} class) for constant-fit trees along with
suitable coercion functions, and \code{"ctree"} and \code{"mob"} discuss the
new \fct{ctree} and \fct{mob} implementations, respectively. Each of the
vignettes can be viewed within \proglang{R} via
\code{vignette(}\emph{``name''}\code{, package = "partykit")} and the underlying
source code (in \proglang{R} with {\LaTeX} text) is also available in the source
package.

@@

The \pkg{partykit} package provides functionality as different levels. First,
there is basic infrastructure for representing, modifying, and displaying trees
and recursive partitions and these tools are mostly intended for developers
and described in the next section. Second, there are tools for inferring trees
from data or for porting trees inferred by other software into \pkg{partykit}.

@@

While originally an important goal for the development of \pkg{partykit} was to
provide infrastructure for the authors' own tree induction algorithms
CTree \citep{Hothorn+Hornik+Zeileis:2006} and MOB \citep{Zeileis+Hothorn+Hornik:2008},
the design was very careful to separate as much functionality as possible into
more general classes that are useful for a far broader class of tree algorithms.
Specifically, to be able to print/plot/predict different trees in a unified way,
there are so-called coercion functions for transforming trees learned in other
software packages (inside and outside of \proglang{R}) into the classes provided
by \pkg{partykit}. Specifically, tree objects learned by \fct{rpart} from the
package of the same name (\citealp{rpart}, implementing CART, \citealp{Breiman+Friedman+Olshen:1984})
and by \fct{J48} from \pkg{RWeka} (\citealp{RWeka}, interfacing \pkg{Weka}'s J4.8
algorithm for C4.5, \citealp{Quinlan:1993}) can be coerced by \fct{as.party}
to the same object class \class{constparty}. This is a general class that can
in principle represent all the major classical tree algorithms with constant
fits in the terminal nodes. Also, the same class is employed for conditional
inference trees (CTree) that can be learned with the \fct{ctree} function
directly within \pkg{partykit} or evolutionary learned trees from \pkg{evtree}
\citep{evtree}.

But not only trees learned within \proglang{R} can be transformed to the
proposed infrastructure but also trees from other software packages. Either
a dedicated interface has to be created using the building blocks described
in the next section (e.g., as done for the J4.8 tree in \pkg{RWeka}) or
PMML (Predictive Model Markup Language) can be used as an intermediate
exchange format. This is an XML standard created by an international
consortium \citep{DMG:2014} that includes a \code{<TreeModel>} tag with
support for constant-fit classification and regression trees. The function
\fct{pmmlTreeModel} allows to read these files and represents them as
\class{simpleparty} objects in \pkg{partykit}. The reason for not using the
\class{constparty} class as above is that the PMML format only stores
point predictions (e.g., a mean or proportion) and not all observations
from the learning sample. So far, the PMML interface has been tested with
output from the \proglang{R} package \pkg{pmml} and \proglang{SPSS}'s
\pkg{AnswerTree} model. The latter includes an implementation of the
QUEST algorithm \citep{Loh+Shih:1997}.

@@

All of these different trees (see Table~\ref{tab:overview} for an overview)
use the same infrastructure at the core but possibly with different options
enabled. In all cases, the functions \fct{print}, \fct{plot}, and \fct{predict}
can be used to create textual and graphical displays of the tree and for
computing predictions on new data, respectively. As an example for the
visualizations, Figure~\ref{fig:ttnc} shows two different trees fitted
to the well-known data on survival of passengers on the ill-fated maiden
voyage of the MS~Titanic: The left panel shows a CART tree with constant fits
learned by \pkg{rpart} and converted to \pkg{partykit}. The left panel
shows a MOB tree learned within \pkg{partykit} with a logistic regression
model for treatment effects in the terminal nodes. Additionally, the
are further utility functions, e.g.,
\fct{nodeapply} can be employed to access further information
stored in the nodes of a tree and \fct{nodeprune} can prune selected nodes.

>>>>

%% software markup
\let\proglang=\textsf
\let\pkg=\textit
\let\code=\texttt
\newcommand{\squote}[1]{`{#1}'}
\newcommand{\dquote}[1]{``{#1}''}
\newcommand{\fct}[1]{\texttt{#1}}
\newcommand{\class}[1]{\squote{\texttt{#1}}}

@@

\begin{abstract}
  The \proglang{R} package \pkg{partykit} provides a flexible toolkit for
  learning, representing, summarizing, and visualizing a wide range of tree-structured
  regression and classification models. The functionality encompasses: (a)~Basic
  infrastructure for \emph{representing} trees (inferred by any algorithm) so that
  unified \code{print}/\code{plot}/\code{predict} methods are available.
  (b)~Dedicated methods for trees with \emph{constant fits} in the leaves (or terminal nodes)
  along with suitable coercion functions to create such trees (e.g., by
  \pkg{rpart}, \pkg{RWeka}, PMML). (c)~A reimplementation of \emph{conditional inference
  trees} (\code{ctree}, originally provided in the \pkg{party} package). (d)~An extended
  reimplementation of \emph{model-based recursive partitioning} (\code{mob}, also originally
  in \pkg{party}) along with dedicated methods for trees with parametric models in the
  leaves. Here, a brief overview of the package and its design is given while more detailed
  discussions of items~(a)--(d) are available in vignettes accompanying the package.
\end{abstract}

@@

In the more than fifty years since \cite{Morgan+Sonquist:1963}
published their seminal paper on ``automatic interaction detection'',
a wide range of methods has been suggested that is usually
termed ``recursive partitioning'' or ``decision trees'' or
``tree(-structured) models'' etc. Particularly influential were the
algorithms CART \citep[classification and regression trees,][]{Breiman+Friedman+Olshen:1984},
C4.5 \citep{Quinlan:1993}, QUEST/GUIDE \citep{Loh+Shih:1997,Loh:2002},
and CTree \citep{Hothorn+Hornik+Zeileis:2006} among many others
\citep[see][for a recent overview]{Loh:2014}. Reflecting the heterogeneity
of conceptual algorithms, a wide range of computational implementations
in various software systems emerged: Typically the original authors of an
algorithm also provide accompanying software but many software systems,
including \pkg{Weka} \citep{Witten+Frank:2005} or \proglang{R} \citep{R},
also provide collections of various types of trees.
Within \proglang{R} the list of prominent packages includes
\pkg{rpart} \citep[implementing CART]{rpart},
\pkg{RWeka} \citep[containing interfaces to J4.8, M5', LMT from \pkg{Weka}]{RWeka}, and
\pkg{party} \citep[implementing CTree and MOB]{party} among
many others. See the CRAN task view ``Machine Learning''
\citep{ctv} for an overview.

@@

The \pkg{partykit} aims at mitigating the latter issue
by providing a common unified infrastructure
for recursive partytioning in the \proglang{R} system for statistical
computing. In particular, \pkg{partykit}
provides tools for representing fitted trees along with printing,
plotting, and computing predictions. The design principles are:
\begin{itemize}
  \item One `agnostic' base class (\class{party}) which can encompass an extremely wide
        range of different types of trees.
  \item Subclasses for important types of trees, e.g., trees with constant
        fits (\class{constparty}) or with parametric models (\class{modelparty})
	in each terminal node (or leaf).
  \item Nodes are recursive objects, i.e., a node can contain child nodes.
  \item Keep the (learning) data out of the recursive node and split structure.
  \item Basic printing, plotting, and predicting for raw node structure.
  \item Customization via suitable panel or panel-generating functions.
  \item Coercion from existing object classes in \proglang{R} (\code{rpart}, \code{J48}, etc.) to the new class.
  \item Usage of simple/fast \proglang{S}3 classes and methods.
\end{itemize}
In addition to all of this generic infrastructure, two specific tree
algorithms are implemented in \pkg{partykit} as well: \fct{ctree}
for conditional inference trees \citep{Hothorn+Hornik+Zeileis:2006}
and \fct{mob} for model-based recursive partitioning \citep{Zeileis+Hothorn+Hornik:2008}.

@@

The \pkg{partykit} package is an add-on package for the \proglang{R} system
for statistical computing. It is available from the Comprehensive \proglang{R}
Archive Network (CRAN) at \url{http://CRAN.R-project.org/package=partykit} and
can be easily installed in the usual ways from within \proglang{R}, e.g., using
\fct{install.packages}. It depends on \proglang{R} (at least 2.15.0) as well
as the base packages \pkg{graphics}, \pkg{grid}, \pkg{stats}, and the recommended
\pkg{survival}. Furthermore, various suggested packages are needed
for certain special functionalities in the package. To install all of these
required and suggested packages in one go, the command
\code{install.packages("partykit", dependencies = TRUE)} can be used.

In addition to the stable release version on CRAN, the
current development release can be installed from \proglang{R}-Forge
\citep{Theussl+Zeileis:2009}. In addition to source and binary packages
the entire version history is available through
\proglang{R}-Forge's \pkg{Subversion} source code management system.

Along with the package extensive documentation with examples is shipped.
The manual pages provide basic technical information on all functions while
much more detailed descriptions along with hands-on examples are provided in
the four package vignettes. First, the vignette \code{"partykit"}
introduces the basic \class{party} class
and associated infrastructure while three further vignettes discuss the tools
built on top of it: \code{"constparty"} covers the eponymous class (as well as
the simplified \class{simpleparty} class) for constant-fit trees along with
suitable coercion functions, and \code{"ctree"} and \code{"mob"} discuss the
new \fct{ctree} and \fct{mob} implementations, respectively. Each of the
vignettes can be viewed within \proglang{R} via
\code{vignette(}\emph{``name''}\code{, package = "partykit")} and the underlying
source code (in \proglang{R} with {\LaTeX} text) is also available in the source
package.

@@

The \pkg{partykit} package provides functionality at different levels. First,
there is basic infrastructure for representing, modifying, and displaying trees
and recursive partitions -- these tools are mostly intended for developers
and described in the next section. Second, there are tools for inferring trees
from data or for importing trees inferred by other software into \pkg{partykit}.

@@

While originally an important goal for the development of \pkg{partykit} was to
provide infrastructure for the authors' own tree induction algorithms
CTree and MOB,
the design was very careful to separate as much functionality as possible into
more general classes that are useful for a far broader class of tree algorithms.
In particular, to be able to print/plot/predict different trees in a unified way,
there are so-called coercion functions for transforming trees learned in other
software packages (inside and outside of \proglang{R}) into the classes provided
by \pkg{partykit}. Specifically, tree objects learned by \fct{rpart} from the
package of the same name (\citealp{rpart}, implementing CART, \citealp{Breiman+Friedman+Olshen:1984})
and by \fct{J48} from \pkg{RWeka} (\citealp{RWeka}, interfacing \pkg{Weka}'s J4.8
algorithm for C4.5, \citealp{Quinlan:1993}) can be coerced by \fct{as.party}
to the same object class \class{constparty}. This is a general class that can
in principle represent all the major classical tree algorithms with constant
fits in the terminal nodes. Also, the same class is employed for conditional
inference trees (CTree) that can be learned with the \fct{ctree} function
directly within \pkg{partykit} or evolutionary trees from package \pkg{evtree}
\citep{evtree}.

Not only trees learned within \proglang{R} can be transformed to the
proposed infrastructure but also trees from other software packages. Either
a dedicated interface has to be created using the building blocks described
in the next section (e.g., as done for the J4.8 tree in \pkg{RWeka}) or
PMML (Predictive Model Markup Language) can be used as an intermediate
exchange format. This is an XML standard created by an international
consortium \citep{DMG:2014} that includes a \code{<TreeModel>} tag with
support for constant-fit classification and regression trees. The function
\fct{pmmlTreeModel} allows to read these files and represents them as
\class{simpleparty} objects in \pkg{partykit}. The reason for not using the
\class{constparty} class as above is that the PMML format only stores
point predictions (e.g., a mean or proportion) rather than all observations
from the learning sample. So far, the PMML interface has been tested with
output from the \proglang{R} package \pkg{pmml} and \proglang{SPSS}'s
\pkg{AnswerTree} model. The latter includes an implementation of the
QUEST algorithm \citep{Loh+Shih:1997}.

@@

All of these different trees (see Table~\ref{tab:overview} for an overview)
use the same infrastructure at the core but possibly with different options
enabled. In all cases, the functions \fct{print}, \fct{plot}, and \fct{predict}
can be used to create textual and graphical displays of the tree and for
computing predictions on new data, respectively. As an example for the
visualizations, Figure~\ref{fig:ttnc} shows two different trees fitted
to the well-known data on survival of passengers on the ill-fated maiden
voyage of the RMS~Titanic: The left panel shows a CART tree with constant fits
learned by \pkg{rpart} and converted to \pkg{partykit}. The right panel
shows a MOB tree learned within \pkg{partykit} with a logistic regression
model for treatment effects in the terminal nodes. Additionally, the
are further utility functions, e.g.,
\fct{nodeapply} can be employed to access further information
stored in the nodes of a tree and \fct{nodeprune} can prune selected nodes.
